---
title: "STAT:5400 Section 3.1 — Random Number Generators"
output: html_document
---

### 2.1 Draw from the uniform distribution

```{r}
(u <- runif(1))
```

### 2.2 Set seed

```{r}
set.seed(5400)
(u <- runif(1))
```

```{r}
set.seed(5400)
length(.Random.seed)
head(.Random.seed)
```

### 2.3 Generate $U \sim \mathrm{Unif}(1, 5)$

*Predict*

```{r}
set.seed(5400)
(u1 <- runif(1, 1, 5))
```

```{r}
set.seed(5400)
(u2 <- runif(1) * 4 + 1)
```

```{r}
help(runif)
```

### 2.5 Generating many uniforms

```{r}
set.seed(5400)
(ulist <- runif(5, 1, 5))
```

```{r}
set.seed(5400)
ulist <- rep(NA, 5)
for (i in seq(5)) {
  ulist[i] <- runif(1, 1, 5)
}
ulist
```

### 2.7 Mean and variance of uniform distribution

```{r}
set.seed(5400)
# generate U1, U2, ..., U100
ulist <- runif(100, 0, 1)
```

```{r}
# population mean is 1/2
mean(ulist)
```

```{r}
# population variance is 1/12 = 0.0833
var(ulist)
```

### 2.10 The probability integral transform, seen

```{r}
par(mfrow = c(1, 2))
X <- runif(1000, 2, 5)
hist(punif(X, 2, 5), main = "Uniform CDF")
X <- rnorm(1000)
hist(pnorm(X), main = "Normal CDF")
```

### 2.12 Q-Q plot (quantile-quantile plot)

```{r}
set.seed(5400)
ulist <- runif(1000, 0, 1)
plot(qunif(ppoints(1000)), sort(ulist),
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles", main = "Q-Q plot")
abline(0, 1)
```

### 3.1 Drawing from Bernoulli distribution

```{r}
# Generate Bern(0.4)
Y <- as.numeric(runif(1) < 0.4)
Y
```

```{r}
set.seed(5400)
# Generate Y1, ..., Y100 ~ Bern(0.4)
Ylist <- as.numeric(runif(100) < 0.4)
```

```{r}
# population mean is 0.4
mean(Ylist)
```

```{r}
# population variance is 0.4 * 0.6 = 0.24
var(Ylist)
```

### 3.2 Drawing from Binomial distribution

```{r}
set.seed(5400)
# Generate X ~ Bin(100, 0.4)
X <- sum(runif(100) < 0.4)
X
```

```{r}
rbinom(1, 100, 0.4)
# rbinom(n, size, prob).
# n is the number of observations.
# size is the number of trials.
```

### 4.2 Drawing from exponential distribution

```{r}
set.seed(5400)
# Generate X1, ..., X1000 ~ exp(2)
Ulist <- runif(1000, 0, 1)
Xlist <- -2 * log(Ulist)
Xlist[1:3]
```

```{r}
mean(Xlist)
var(Xlist)
```

### 4.3 Histogram against the density

*By hand*

```{r}
par(mfrow = c(1, 2))
hist(Xlist)
xs <- seq(0.001, 12, len = 200)
plot(xs, dexp(xs, rate = 1),        # <- change this line
  type = "l", ylim = c(0, 1),
  xlab = "X", main = "exp(2)", ylab = "f(x)")
```

### 4.4 Q-Q plot for the exponential

*By hand*

```{r}
probs <- ppoints(1000)
plot(probs, probs,                  # <- change this line
  xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
  main = "Q-Q plot for exponential distribution")
abline(0, 1)
```

### 4.5 The `rexp` function in R

```{r}
set.seed(5400)
Xlist <- rexp(1000, rate = 1/2)
Xlist[1:3]
```

```{r}
mean(Xlist)
var(Xlist)
```

*Predict*

```{r}
set.seed(5400)
head(-2 * log(runif(3)))
set.seed(5400)
head(rexp(3, rate = 1/2))
```

### 4.6 Drawing from the standard Cauchy distribution

```{r}
set.seed(5400)
# Generate X1, ..., X1000 ~ Cauchy
Ulist <- runif(1000, 0, 1)
Xlist <- tan(pi * (Ulist - 0.5))
```

```{r}
par(mfrow = c(1, 2))
hist(Xlist)
xs <- seq(-10, 10, len = 200)
plot(xs, dcauchy(xs), type = "l", ylim = c(0, 1),
  xlab = "X", main = "PDF of Cauchy distribution", ylab = "f(x)")
```

*By hand*

```{r}
probs <- ppoints(1000)
plot(probs, probs,                  # <- change this line
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles",
  main = "Q-Q plot for Cauchy distribution")
abline(0, 1)
```

### 4.7 Cauchy and $t$

```{r}
set.seed(5400)
mean(rcauchy(10000))
mean(rcauchy(10000))
mean(rcauchy(10000))
```

### 4.8 Sampling the next word

```{r}
words <- c("reject", "accept", "retain")
p <- c(0.665, 0.245, 0.090)

set.seed(5400)
u <- runif(1)
words[findInterval(u, cumsum(p)) + 1]   # one uniform, one word
```

```{r}
# 10,000 draws: the frequencies match p
u <- runif(10000)
table(words[findInterval(u, cumsum(p)) + 1]) / 10000
```

### 5.2 Accept/reject sampling (cont'd)

```{r}
set.seed(5400)
n <- 2000    # total samples
# Each row is a pair of V1 and V2. n rows in total.
dat <- matrix(runif(n * 2, -1, 1), n, 2)
# Accept if V_1^2 + V_2^2 <= 1.
accept <- (dat[, 1]^2 + dat[, 2]^2) <= 1
```

```{r}
mean(accept)   # which is about pi/4
pi / 4
```

```{r}
# Accepted pairs of V1 and V2
V <- dat[accept, ]
# Rejected pairs of V1 and V2
V_out <- dat[!accept, ]
```

```{r}
# draw a circle
theta <- seq(0, 2 * pi, len = 1000)
plot(cbind(sin(theta), cos(theta)),
  type = "l", xlab = "x", ylab = "y")
points(V, pch = 2, col = "blue")      # Accepted pairs
points(V_out, pch = 1, col = "red")   # Rejected pairs
```

```{r}
c_bound <- optimize(function(x) dnorm(x) / dcauchy(x),
                    c(-3, 3), maximum = TRUE)$objective
c_bound
```

```{r}
xs <- seq(-6, 6, length = 400)
plot(xs, c_bound * dcauchy(xs), type = "n",
  xlab = "x", ylab = "density", ylim = c(0, 0.52))
polygon(c(xs, rev(xs)),
  c(c_bound * dcauchy(xs), rev(dnorm(xs))),
  col = "gray90", border = NA)          # rejected
lines(xs, c_bound * dcauchy(xs), lwd = 2, col = "#b8860b")
lines(xs, dnorm(xs), lwd = 2)
legend("topright", bty = "n", lwd = 2,
  col = c("#b8860b", "black"),
  legend = c("c g(x), Cauchy envelope", "f(x), standard normal"))
```

```{r}
set.seed(5400)
n <- 1e4
Y <- rcauchy(n)
U <- runif(n)
keep <- U <= dnorm(Y) / (c_bound * dcauchy(Y))

mean(keep)      # close to 1 / c_bound
1 / c_bound
```

```{r}
X <- Y[keep]
hist(X, breaks = 40, freq = FALSE, main = "", xlab = "x")
curve(dnorm(x), add = TRUE, lwd = 2)
```

### 6.2 Polar method for the standard normal

```{r}
# an OK code
Rsq <- V[, 1]^2 + V[, 2]^2
m <- sqrt(-(2 * log(Rsq)) / (Rsq))
X <- m * V
```

```{r}
# a better code
Rsq <- V[, 1]^2 + V[, 2]^2
Rsq <- Rsq + .Machine$double.xmin
m <- sqrt(-(2 * log(Rsq)) / (Rsq))
X <- m * V
```

### 6.3 Checking the normals

```{r}
par(mfrow = c(2, 2))
# histograms
hist(X[, 1], main = "Histogram of X1")
hist(X[, 2], main = "Histogram of X2")
# Q-Q plots
probs <- ppoints(NROW(X))
plot(qnorm(probs), sort(X[, 1]),
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles", main = "Q-Q plot for X1")
abline(0, 1)
plot(qnorm(probs), sort(X[, 2]),
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles", main = "Q-Q plot for X2")
abline(0, 1)
```

### 6.4 Q-Q plot for $\mathrm{N}(\mu, \sigma^2)$

```{r}
newX <- X[, 1] * 2 + 4
dat_quant <- sort(newX)
thr_quant <- qnorm(ppoints(length(newX)))
plot(thr_quant, dat_quant,
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles",
  main = "Q-Q plot for X ~ N(4, 2)")
abline(a = 4, b = 2) # a = intercept; b = slope
```

### 6.5 `qqnorm` and `qqline`

```{r}
qqnorm(newX)
qqline(newX)
```

### 6.6 What R actually uses

```{r}
set.seed(5400)
qnorm(runif(1))
```

```{r}
set.seed(5400)
rnorm(1)
```

### 6.7 Standard Cauchy, again

```{r}
Y <- V[, 1] / V[, 2]
summary(Y)
```

### 6.8 $\chi_v^2$ distribution

```{r}
set.seed(5400)
Zlist <- rnorm(10)
Y <- sum(Zlist^2)
Y
```

### 6.9 Student's $t$ distribution

```{r}
curve(dnorm(x, 0, 1), xlim = c(-3, 3),
  main = "Normal and t",
  xlab = "X", ylab = "f(x)", lwd = 3, lty = 1)
curve(dt(x, 1), xlim = c(-3, 3), add = TRUE, lty = 2, col = "red")
curve(dt(x, 2), xlim = c(-3, 3), add = TRUE, lty = 3, col = "blue")
curve(dt(x, 5), xlim = c(-3, 3), add = TRUE, lty = 4, col = "green")
abline(h = 0)
legend("topright",
  c("normal", "t(1)", "t(2)", "t(5)"),
  lty = 1:4,
  col = c("black", "red", "blue", "green"))
```

### 7.4 Generate one copy from $\mathrm{N}_p(\boldsymbol{\mu}, \boldsymbol{\Sigma})$

```{r}
set.seed(5400)
# sample size and dimension
n <- 1000
p <- 5
# var-covariance matrix
Sigma <- matrix(0.5, p, p)
diag(Sigma) <- 1
mu <- seq(p)   # mean vector
```

```{r}
# eigen-decomposition
eig <- eigen(Sigma)
Sigma.sqrt <- eig$vectors %*%
  tcrossprod(diag(sqrt(eig$values)), eig$vectors)
```

```{r}
Z <- rnorm(p)
X <- mu + Sigma.sqrt %*% Z
X
```

### 7.5 Generate multiple copies from $\mathrm{N}_p(\boldsymbol{\mu}, \boldsymbol{\Sigma})$

```{r}
set.seed(5400)
# each row of Z is from N(0, 1)
Z <- matrix(rnorm(n * p), n, p)
# each row of X is from N(mu, Sigma)
X <- tcrossprod(rep(1, n), mu) + Z %*% Sigma.sqrt
```

```{r}
(Xbar <- colMeans(X))
```

```{r}
(S <- cov(X))
```

### 7.6 R library `mvtnorm`

```{r}
set.seed(5400)
Z <- rnorm(p)
X <- mu + Sigma.sqrt %*% Z
X[1:2]
```

```{r}
set.seed(5400)
library(mvtnorm)
X <- rmvnorm(n = 1, mean = mu, sigma = Sigma)
X[1:2]
```
